---
title: "PS2010 Code Bank"
author: ""
date: ""
output:
  html_document:
    theme: flatly
    highlight: haddock
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

Welcome to the PS2010 Code Bank. Here you will find a collection of the R code used for all data analyses covered across the module. Use the menu on the left hand side to navigate.

# **Basics**

## Install and Load Packages

***

### **Install a package**

```{r xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\" style=\"font-size: 24px;\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D; font-size: 24px;\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r, eval=FALSE}
install.packages(NULL)
install.packages("tidyverse")
```
Change `NULL` to the name of the package.

***

### **Load a package**
```{r, eval=FALSE}
library(NULL)
library(tidyverse)
```
Change `NULL` to the name of the package.

***

## Importing Data

### **Set the Working Directory**
```{r, eval=FALSE}
setwd(NULL)
setwd(C:/Users/username/documents)
```
Change `NULL` to the folder path.

Remember you can also set the working directory using the menu options: 

Session -> Set Working Directory -> Choose Directory

***

### **Check the Working Directory**
```{r, eval=FALSE}
getwd()
```
This will tell you the current working directory.

***

### **Import Data**
```{r, eval=FALSE}
mydata = read_csv(NULL)
mydata = read_csv("experiment_data.csv")
```
`mydata` is the object that we save the data under. You can call it whatever you like.

Change `NULL` to the name of the file. 

Remember to add ".csv" at the end of the filename. 

R only reads .csv data files.

Note: `read_csv()` requires the tidyverse package.

***

# **Describing Data**

# **T-tests**

# **One-Way ANOVA**

# **Factorial ANOVA**

