---
title: "PS2010 Code Bank"
author: ""
date: ""
output:
  html_document:
    theme: flatly
    highlight: haddock
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

Welcome to the PS2010 Code Bank. Here you will find a collection of the R code used for all data analyses covered across the module. Use the menu on the left hand side to navigate. 

Some important tips:

* The data used in all of this code will be referred to as "mydata". Make sure you change this if you have called your data something else.

* Where you see `NULL`, `NULL1`, `NULL2` etc is where you will need to add information based on the data set you are using.

# **1) Basics**

## Install and Load Packages

***

### **Install a package**

```{r xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\" style=\"font-size: 24px;\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D; font-size: 24px;\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r, eval=FALSE}
install.packages(NULL)
install.packages("tidyverse") #An example of this code
```
Change `NULL` to the name of the package.

***

### **Load a package**
```{r, eval=FALSE}
library(NULL)
library(tidyverse) #An example of this code
```
Change `NULL` to the name of the package.

***

## Importing Data

### **Set the Working Directory**
```{r, eval=FALSE}
setwd(NULL)
setwd(C:/Users/username/documents) #An example of this code
```
Change `NULL` to the folder path.

Remember you can also set the working directory using the menu options: 

Session -> Set Working Directory -> Choose Directory

***

### **Check the Working Directory**
```{r, eval=FALSE}
getwd()
```
This will tell you the current working directory.

***

### **Import Data**
```{r, eval=FALSE}
mydata = read_csv(NULL)
mydata = read_csv("experiment_data.csv") #An example of this code
```
`mydata` is the object that we save the data under. You can call it whatever you like.

You can see the data in the Environment panel (top right in RStudio).

Change `NULL` to the name of the file. 

Remember to add ".csv" at the end of the file name. 

R only reads .csv data files.

Note: `read_csv()` requires the tidyverse package.

***

# **2) Describing Data**

## Summary Statistics

***

### Quickly Summarising a Data Set
```{r, eval=FALSE}
summary(NULL)
summary(mydata) #An example of this code
```
Change `NULL` to the name of the object/data file you want to summarise.

***

### Splitting by Group or Variable
```{r, eval=FALSE}
NULL1 = mydata |>
  group_by(NULL2) |>
  summarise(
    mean = mean(NULL3)
    sd = sd(NULL3)
  )
```
Change `NULL1` to the name of the object/data file you want to create (i.e that will hold the descriptive statistics)

Change `NULL2` to the name of the variable you want to split the data by.

Change `NULL3` to the name of the dependent variable you want the descriptive statistics for.

See the example below:

```{r, eval=FALSE}
summary_by_group = mydata |>
  group_by(drink) |>
  summarise(
    mean = mean(bpm)
    sd = sd(bpm)
  )
```
Line by line analysis:

1. The first line creates an object called `summary_by_group` which uses `mydata` AND THEN

2. It will group `mydata` by the grouping variable, in this example called `drink` AND THEN

3. It asks for the `summarise()` function.

4. It will call the mean `mean` and ask for the mean of our dependent variable using `mean(bpm)`

5. It will call the standard deviation `sd` and ask for the standard deviation of our dependent variable using `sd(bpm)`

6. It will close the code with a final `)`

***

### Mean and Standard Deviation for an Entire Data Set

Use the code above again, but just remove the `group_by()` line of code. See example below:

```{r, eval=FALSE}
summary_by_group = mydata |>
  summarise(
    mean = mean(bpm)
    sd = sd(bpm)
  )
```

***

# **3) T-tests**

# **4) One-Way ANOVA**

# **5) Factorial ANOVA**

